---
title: "Linear regression in R & Python"
subtitle: |
  | Estimating the coefficients in R;
  | Model analytics and diagnostics 
author: 
  - Sonja Petrovic^[Sonja Petrović,  Associate Professor of Applied  Mathematics, College of Computing, Illinios Tech.  [Homepage](https://www.SonjaPetrovicStats.com), [Email](mailto:sonja.petrovic@iit.edu).] 
date: "Week 12, ITMD/ITMS/STAT 514, Sp21"
output: rmdformats::readthedown
urlcolor: darkblue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '' #, fig.width = 3, fig.height = 3
  )
```


--- 

# Background 

> What is the purpose of these notes? 

1. Provide an example of simple and multiple linear regression; 
2. Learn the linear regression function in `R`'
3. Learn the linear regression function in `Python`. 

> What is the format of this document?

This document  was created using `R Markdown`. You can read more about it [here](https://www.markdownguide.org/getting-started/) and check out a [cheat sheet here](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf), which will guide you through installing RStudio, and from there the moment you create  a new `.Rmd` document, it will be a  working template to start from.  If you are used to using `LaTeX`, no worries: it can be embedded into `Markdown`, with overall simpler formatting. I hope you find this useful! 

### Installing and loading packages

Just like every other programming language you may be familiar with, R's capabilities can be greatly extended by installing additional "packages" and "libraries".

To **install** a package, use the `install.packages()` command.  You'll want to run the following commands to get the necessary packages for today's lab:

```
install.packages("rmdformats")
install.packages("ggplot2")
install.packages("knitr")
install.packages("ISLR")
```

You only need to install packages once.  Once they're installed, you may use them by **loading** the libraries using the `library()` command.  For today's lab, you'll want to run the following code

```{r}
library(ggplot2) # graphics library
library(knitr)   # contains kable() function
options(scipen = 4)  # Suppresses scientific notation

library("ISLR") # to obtain the data sets used in the book 
library("reticulate") # so we can use Python
```

```{r}
# this is another r-code chunk to install the packages i need for Python: 
py_install("pandas")
py_install("statsmodels")
py_install("matplotlib")
```


# Context 

Recall some important questions about linear regression model

1. Is at least one of the predictors $X_1, \dots, X_p$ useful in predicting the response?
2. Do all the predictors help to explain Y, or is only a subset of the predictors useful?
3. How well does the model fit the data?
4. Given a set of predictor values, what response value should we predict,
and how accurate is our prediction?

## Review: 

These concepts will be important to keep in mind: 
* what is the **population regression line**?
* what is the **estimated regression line**? 
* what is the meaning of the **estimated intercept** when you estimate the regression line?
* what is the meaning of the **estimated coefficients**?


# Simple linear regression example in `R`

Throughout this handout, we will be working on the `Auto` data set that comes with the `ISLR` package in `R`. 
```{r}
require(ISLR)
data(Auto)
```

## `lm` and output

Here is how to do simple linear regression in `R`. 

* Use the `lm()` function to perform a simple linear regression with `mpg` as the response and `horsepower` as the predictor using the following code . 
```{r}
fit.lm <- lm(mpg ~ horsepower, data=Auto)
```

Wait a minute.  Where is the output?? 

> As usual, it's stored in the object to the left of `<-` and so it is not printed. 

 Let's take a look at the `fit.lm` object. 
Use the `summary()` function to print the results. 
```{r}
summary(fit.lm)
```


## Interpreting the output

*  Is there a relationship between the predictor and the response?

>  Yes. 

* How strong is the relationship between the predictor and the response?

> $p$-value is close to 0: relationship is strong

* Is the relationship between the predictor and the response positive or negative?

>  Coefficient is negative: relationship is negative 


* What is the predicted `mpg` associated with a `horsepower` of $98$? What are the associated 95% confidence and prediction intervals?

First you have to make a new data frame object which will contain the new point: 
```{r}
new <- data.frame(horsepower = 98)
predict(fit.lm, new) # predicted mpg
predict(fit.lm, new, interval="confidence") # conf interval
predict(fit.lm, new, interval="prediction") # pred interval
```

   * $\rightarrow$ confidence interval vs. prediction interval $\leftarrow$ 

>  What is the difference between confidence and prediction intervals!?   $\rightarrow$ we will learn in the next lecture!! 


## Simple plots


Plot the response and the predictor. Use the `abline()` function to display the least squares regression line.

```{r} 
plot(Auto$horsepower, Auto$mpg)
abline(fit.lm, col="red")
```


Use the `plot()` function to produce diagnostic plots of the least squares regression fit. Comment on any problems you see with the fit.

```{r} 
par(mfrow=c(2,2))
plot(fit.lm)
```

* residuals vs fitted plot shows that the relationship is non-linear


# Simple linear regression example in `Python`

> Please note: 
 In this section, all code chunks are `python` code chunks. This means we have written `{python}` rather than `{r}` at the top. 


```{python}
import pandas as pd
import statsmodels.api as sm
import matplotlib.pyplot as plt
```

```{python, include = FALSE}
path_to_file="/Users/sonja/Documents/GitHub/StatITM514fa20/Stat514sp21/data/auto.csv"
# path_to_file="Desktop/auto.csv"
df= pd.read_csv(path_to_file)  # Remember to change the path of the file when you are running the code and use thep ath to file you downloaded to your computer 
```


```
path_to_file="Desktop/auto.csv"
df= pd.read_csv(path_to_file)  # Remember to change the path of the file when you are running the code and use thep ath to file you downloaded to your computer 
```

Now that we have imported the data let's take a peek:
```{python}
df.head()
```


## `sm.OLS` and output
The target variable that we are interested to study is "mpg" and we want to see the relationship of it with other variables.

First, a simple linear regression on "horsepower" and "mpg" is done like this: 

```{python}
X,y= df["horsepower"] , df["mpg"]
```

For this purpose we are using `statsmodels.api`. One thing to note before making the model is to keep this in mind that for containing an intercept in our model we should add a new `"constant variable"` to our predictor(s) because `statsmodels.api` object does perform a regression without an intercept! 


```{python}
#Adding constant because we want the intercept
X = sm.add_constant(X)

#with the following command you can fit the model, remember to put the y(target) first!
model =sm.OLS(y,X).fit()
```

```{python}
#The following command will provide a good summary of the fitted model:
model.summary()
```

## Simple plots 

Now lets plot the fitted model: 
```{python}
fix, ax = plt.subplots()

ax.scatter(df["horsepower"], model.predict(), label='fitted')
ax.scatter(df["horsepower"], df["mpg"], label='original')

ax.legend()
ax.set_title("Linear model fitted values vs the original dataset")
ax.set_xlabel("horsepower")
ax.set_ylabel("mpg")
plt.show()
```


## Prediction

If you want to predict a new value such as `98`, you have to give it as an input like `[1,98]` for predicting because of the intercept, if you want to predict many values like a set called New_values, all you need to put in the argument is : `exog=[1,New_vallues]`. 
```{python}
model.predict(exog=[1,98])
```


# Quick reference to multiple linear regression


Q: is there a relationship between the Response and Predictor? 

* Multiple case:  $p$ predictors;  we need to ask whether all of the regression coefficients are zero: $\beta_1=\dots=\beta_p=0$? 
  * Hypothesis test: $H_0:\beta_1=\dots=\beta_p=0$ vs. $H_1:$ at least one $\beta_i\neq 0$. 
  * Which statistic? \[F= \frac{(TSS-RSS)/p}{RSS/(n-p-1)}.\]
    * TSS and RSS defined as in simple case. 

* when there is no relationship between the response and predictors, one would expect the F-statistic to take on a value close to 1.   [this can be proved via expected values]
* else $>1$. 

* $\rightarrow$ in case of large $p$, may want to measure *partial effects*, and do some *variable selection* 


# Multiple linear regression  example in  `R`


* Produce a scatterplot matrix which includes all of the variables in the data set.
```{r}
pairs(Auto,lower.panel = NULL)
```
What are we looking at? (Read the help page on `pairs`.)

* Compute the matrix of correlations between the variables using the function `cor()`. You will need to exclude the `name` variable, which is qualitative.

```{r}
cor(subset(Auto, select=-name))
```

*  Use the `lm()` function to perform a multiple linear regression with `mpg` as the response and *all* other variables except name as the predictors. Print results. 
 
```{r}
fit.lm <- lm(mpg~.-name, data=Auto)
summary(fit.lm)
```
Let's refer to the output in this last code chunk: 

* Is there a relationship between the predictors and the response?

   - There is a relationship between predictors and response. 

* Which predictors appear to have a statistically significant relationship to the response?

   - `weight`, `year`, `origin` and `displacement` have statistically significant relationships 

* What does the coefficient for the `year` variable suggest?

   - $0.75$ coefficient for `year` suggests that later model year cars have better (higher) `mpg`. 


* Use the `plot() function to produce diagnostic plots of the linear regression fit. 
```{r}
par(mfrow=c(2,2))
plot(fit.lm)
```
* Comment on any problems you see with the fit. Do the residual plots suggest any unusually large outliers? 
  * There is evidence of non-linearity

  

# Multiple linear regression  example in `Python` 

```{python}
#we are going to use all the features except 'name' and of course dropping the 'mpg' too
X = df.drop(labels=['name','mpg'],axis=1)  
y=df['mpg']
```

```{python}
X = sm.add_constant(X)  #adding constant for intercept
model =sm.OLS(y,X).fit()
```

```{python}
model.summary()
```




# Appendix
Some resources and further reading: 

* Read about model diagnostics in [ISLR book](https://www.statlearning.com/). 
* Read about regression (more in depth) in [ISLR book](https://www.statlearning.com/).


* To view `lm {stats}	R Documentation`, type: 
```
help(lm)
```
and/or visit:  
<https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm> 



# License

This document  is created for Math 514, Spring 2021, at Illinois Tech. 
While the course materials are generally not  to be distributed outside the course without permission of the instructor,  this particular set of notes is licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).



Contents of this lecture is based on the chapter 3 of the textbook Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani, 
*' An Introduction to Statistical Learning: with Applications in R'*. 

Python code for this lecture is developed by Amirreza Eshraghi, our TA in 2020/21. 

[![Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License][image]][hyperlink]

  [hyperlink]: https://creativecommons.org/licenses/by-nc-sa/4.0/
  [image]: https://i.creativecommons.org/l/by-nc-sa/4.0/88x31.png
    


