---
title: "Topic 1.1 - Descriptive statistics" 
author: 
  - Sonja Petrovic^[Sonja Petrović,  Associate Professor of Applied  Mathematics, College of Computing, Illinios Tech.  [Homepage](https://www.SonjaPetrovicStats.com), [Email](mailto:sonja.petrovic@iit.edu).] 
date: "Week 2, ITMD/ITMS/STAT 514, Sp21"
output: rmdformats::readthedown
urlcolor: darkblue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = ''
  )
```


--- 

# Background 

> What is the purpose of these notes? 

1. Provide a few small examples of descriptive statistics;
2. See bar plots, histograms, boxplots;
3. Frequency distribution - visualization. 

> What is the format of this document?

*(One more time, I will add these notes.)*  
This document  was created using `R Markdown`. You can read more about it [here](https://www.markdownguide.org/getting-started/) and check out a [cheat sheet here](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf), which will guide you through installing RStudio, and from there the moment you create  a new `.Rmd` document, it will be a  working template to start from.  If you are used to using `LaTeX`, no worries: it can be embedded into `Markdown`, with overall simpler formatting. I hope you find this useful! 

# Loading required packages

To run some of the commands below, you need to load `qqplot2`: 
```{r}
library(ggplot2)
```

# Example of a small data set
In R, the letter "c" stands for "column" and is used to create a data vector (column vector) of numbers. Think of it as an ordered list, simply. 

## Example from a book on car battery data
Note that we can manually enter data, in a list, and store it under the name  `car.batteries`. (Of course this is *not* how you'll be loading the data in practice!)
```{r carbatteries}
car.batteries<-c(
  2.2,4.1,3.5, 4.5, 3.2, 3.7, 3.0, 2.6, 3.4, 1.6, 3.1, 3.3, 3.8, 3.1, 4.7, 3.7, 2.5, 
  4.3, 3.4, 3.6, 2.9, 3.3, 3.9, 3.1, 3.3, 3.1, 3.7, 4.4, 3.2, 4.1, 1.9, 3.4, 4.7, 
  3.8, 3.2, 2.6, 3.9, 3.0, 4.2, 3.5
)
```
Let's plot the histogram of the above! 
```{r fig.width=5}
hist(car.batteries)
hist(car.batteries,freq=FALSE)
hist(car.batteries,freq=FALSE,ylim = c(0,1),xlim=c(1,6),main = paste("Histogram of..."),col="blue")
lines(density(car.batteries),col="red") # estimated distribution 
```

Note that we can change the options... study the code. Note the figure width can be controlled, as well as the limits on the x and y axes, the main title, and the color. 

# Using descriptive statistics to classify 'shape of data'

As n grows, the histograms may change. 
Here are examples of various normal distributions (you don't need to know what that _really_ means yet). 
```{r descriptivestats}
# set sample size so the code below can be re-run by just changing this quantity: 
sample.size = 100
```
 
 What descriptive statistics would you use to summarize this kind of data? 
```{r descriptivestats.cont}
par(mfrow = c(1,3)) # aligns figures in a grid! in this case, a 1x3 grid. 
# sample the data: 
x <- rnorm(n=sample.size)
y <- rnorm(n=sample.size,mean=250,sd=300)
z <- rnorm(n=sample.size,mean=10,sd=1)
# plot the histograms: 
hist(x)
hist(y)
hist(z)
```

Let us introduce the "five-number summary" + mean: 
```{r fivenumsummary}
summary(x)
summary(y)
summary(z)
```

What happens when we change the sample size to $n=1000$ and re-run the same commands as above? 
```{r rerundescriptive, echo=FALSE}
sample.size = 1000
par(mfrow = c(1, 3))
x <- rnorm(n=sample.size)
hist(x)
y <- rnorm(n=sample.size,mean=250,sd=300)
hist(y)
z <- rnorm(n=sample.size,mean=10,sd=1)
hist(z)
```

Let us visualize the quantile summaries of a  simple data set:
```{r quantile.visualization}
my.data<-x
fivenum(my.data)
par(mfrow = c(1, 2))
boxplot(my.data)
boxplot(my.data)
abline(h = min(my.data), col = "Blue")
abline(h = max(my.data), col = "Yellow")
abline(h = median(my.data), col = "Green")
abline(h = quantile(my.data, c(0.25, 0.75)), col = "Red")
```
In the code above, the third line par(mfrow = c(1, 2)) creates a grid of size 1x2 for plots; it divides the plot area into a grid so you see several plots on the same page as opposed to separately. Try changing the 1 and the 2 to something else! 



## Online resources that are extremely useful:
http://www.stat.cmu.edu/~cshalizi/rmarkdown/ 
and 
https://anaconda.org/anaconda/markdown



# License

This document  is created for Math 514, Spring 2021, at Illinois Tech. 
While the course materials are generally not  to be distributed outside the course without permission of the instructor,  this particular set of notes is licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).


[![Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License][image]][hyperlink]

  [hyperlink]: https://creativecommons.org/licenses/by-nc-sa/4.0/
  [image]: https://i.creativecommons.org/l/by-nc-sa/4.0/88x31.png
    

