---
title: "Homework, worksheet, and lab template"
author: "Your Name Here"
date: Prepared for  ITMD/ITMS/STAT 514, Spring 2021
output: html_document
urlcolor: magenta
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.width = 3, fig.height = 3
  )
```


# Setup


### Changing the author field and file name.

* (a) Change the `author:` field on the Rmd document from Your Name Here to your own name.

* (b) Rename this file to "hwX_YourHameHere.Rmd", where YourNameHere is changed to your own name.

# Cleanup 
Once you are done with this setup, updating the author and date fields, **delete** the *Setup* and the *Cleanup* sections of this document! 


##  Packages

```{r message=FALSE}
# add packages you need for this assignment
library(tidyverse) 
```


----- 


# Preparation (in-class and group work)

 We talked about basic results on sampling distributions: for a mean of a population and a difference in means in two populations. We've also gone over some basic R/Python functionality within RStudio, basic functions, packages, etc. You can use all of that knowledge in this homework set. **UPDATE ME**

**Pro-tip**: Working through these examples, do reflect on the question of 'what is the cleanest shortest way I can call a function in R/Python to compute what I need?' Chances are, we've covered that function/code in the lecture examples and worksheets! 

**... And what's next?** After this, we will go deeper into concepts needed for statistical inference beyond the probability comptuations we are doing right now. **UPDATE ME**



## How to work with this document

There are several ways to format your answer, but whatever you do, please make sure it is *readable* by a human and clean. That is, don't leave stray comments and commented instructions in your submitted work. You may use a quote code `>` to start your answer after any r/python code chunks you are typing. An example is provided in the first question below. 

----- 

# Part ONE: Clarifying the concepts
In these four problems, you will ..... 

## 1. Sampling distribution of the sample mean
### Instructions
_Review the theory about ...._

*Justify both your computation and the reasoning behind your text  answer. Include the plot in your answer and shade the area that represents the probability you computed.* 


### Task
problem statmement 

(a) question a.

> [optional] Type your explanation here of how and why you may be running any code below. 

```{r}
# type the code you need to compute the answer here.
# silly example I will use below: 
my.output <- 2+3 
```

> Type your explanation here and the answer. Make sure to use **in-line code chunks** for any values you reference from the above computation. Example: the output of my computation above is `r my.output`. 

(b) question b. 

> [optional] Type your explanation here of how and why you may be running any code below. 

```{r}
# type the code you need to compute the answer here.
# silly example I will use below: 
my.output <- 2+3 
```

> Type your explanation here and the answer. Make sure to use **in-line code chunks** for any values you reference from the above computation. Example: the output of my computation above is `r my.output`. 
