---
title: "Topic 2: Sampling distribution of the difference of 2 means"
subtitle: |
  | Background for inference of location parameter in location/scale families
author: |
  | Sonja Petrović
  | Created for ITMD/ITMS/STAT 514
date: "Spring 2021." 
urlcolor: darkblue
output: 
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "beaver"
    includes:
      in_header: header_beamer.tex
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.align = 'center'  , out.width="40%"
)
knitr::include_graphics
```



# Sampling distributions

\hfill\emph{(Blue text = links!)}

\vfill 

**Review:** 

(1) How does inference relate to analytics? 

(2) What is  a  sampling distribution? 

(3) Sampling distribution of the mean: in previous lecture, we applied the Central Limit Theorem to the one-paramter location problem. 

# Sampling distribution of the difference in means

\textcolor{myred}{Setup}

- Previous  case study:   
  - statistical inference about a single population mean $\mu$.
- What if you have a comparative  experiment in which two methods are compared? 
  - Two 'methods'? $\rightarrow$ manufacturing methods;  statistical/machine learnig methods; etc. 
  $$\mu_1 \mbox{ VS. } \mu_2$$
... compare... how? 
$\mu_1 ?= \mu_2$? \pause Does $\mu_1-\mu_2=0$? 
\pause 

- Compare the means of the two populations (one representing  each method), denoted by $\mu_1$ and $\mu_2$, or sometimes  $\mu_A$ and $\mu_B$. 

\redblock{{Underlying question:} 
What is the  sampling distribution of $\mu_1-\mu_2$?? 
}


# Theoretical distribution 

Suppose that two independent samples of size $n_1$ and $n_2$   are drawn at random from two populations, discrete or continuous, with means  $\mu_1$ and $\mu_2$ and  variances $\sigma_1$ and $\sigma_2$, respectively. Then: 

\redblock{{Theorem}
  The sampling distribution of the differences of means, $\Xbar_1-\Xbar_2$, is approximately normally distributed with: 
  \[ \attn{\mu_{\Xbar_1-\Xbar_2}} =  \mu_1-\mu_2, \quad 
   \attn{\sigma^2_{\Xbar_1-\Xbar_2}} = \frac{\sigma^2_1}{n_1}+\frac{\sigma^2_2}{n_2}\]

Hence, 
\[
  Z = \frac{(\Xbar_1-\Xbar_2)-( \mu_1-\mu_2)}{\sqrt{\sigma^2_1/n_1+\sigma^2_2/n_2}} 
\]
is approximatley a standard normal random variable. 
}


# Exploring the meaning of the theorem: R

Suppose we just take one sample each, $X$ and $Y$, of size $100$ from two populations like this: 
```{r exploreDiffMeansIllustate}
  x <- rnorm(n=100,mean=25,sd=10)
  Xbar <- mean(x)
  y <- rnorm(n=100,mean=35,sd=10)
  Ybar <- mean(y) 
  Xbar-Ybar 
```
Of course this is just *one* value of the staistic $\Xbar-\overline{Y}$.... Now, repeat!  

---

Values  of $\Xbar-\overline{Y}$  in  repeated sampling, $n_1=n_2=n$: 

```{r exploreDiffMeans,  echo=TRUE}
sample.size=100
diff.in.means <- replicate(10, 
                           mean(rnorm(n=sample.size,mean=25,sd=10))-
                             mean(rnorm(n=100,mean=35,sd=10)))
```

```{r exploreDiffMeans-2, fig.height=4, fig.width=4, echo=FALSE}
par(mfrow = c(1, 2))
hist(diff.in.means,main="n=100, 10 reps",xlab="Difference of two means")
diff.in.means <- replicate(1000, 
       mean(rnorm(n=sample.size,mean=25,sd=10))
       - mean(rnorm(n=100,mean=35,sd=10)))
hist(diff.in.means,main="n=100, 1000 reps",xlab="Difference of two means")
```


\redblock{{Aha!} Are you able to infer anything from the histograms? }


# Exploring the meaning of the theorem: Python


Packages needed: 

```{r, include=TRUE}
  library(reticulate)
  scipy <- import("scipy")
  numpy <- import("numpy")
```

Next slide = code to obtain **one** value of $\overline X - \overline Y$. 

--- 

**Python code**
```{python}
from scipy.stats import norm
import numpy as np  
x = norm.rvs(loc=25,scale=10,size=100)
Xbar = np.mean(x)
y = norm.rvs(loc=35,scale=10,size=100)
Ybar = np.mean(y) 
Xbar-Ybar 
```

#  Case  Study 2: paint drying time

**Problem** 
Two independent experiments are run in which two different
types of paint are compared. 18 specimens are painted using type A, and
the drying time, in hours, is recorded for each. The same is done with type B.
The population standard deviations are both known to be 1.0.

\redblock{{Question:}
Assuming that the mean drying time is equal for the two types of paint, \attn{find
$P(\Xbar_A-\Xbar_B>1)$}, where $\Xbar_A$ and $\Xbar_B$ are average drying times for samples of size
$n_A=n_B=18$.  
}



# Case study 2: Solution - code 

The  sampling distribution of $\Xbar_A-\Xbar_B$  is ...\pause
approximately normal, mean $\mu_{\Xbar_A-\Xbar_B}=\mu_A-\mu_B=0$  and variance $\sigma^2_{\Xbar_A-\Xbar_B} = 1/18+1/18=1/9$. 


```{r, include=FALSE}
  library(tigerstats)
```

```{r exPaintSoln, echo=FALSE,  fig.width = 2.5, fig.height = 2.5}
pnormGC(1, region="above", mean=0, sd=sqrt(1/9),graph=TRUE)
```

# Case study 2: Solution - compute 

The probability that we   compute is given by:
\[ 
P(\Xbar_A-\Xbar_B> 1) = P\left(\frac{\Xbar_A-\Xbar_B-0}{\sqrt{1/9}}\geq \frac{1-0}{\sqrt{1/9}}\right)=P(Z>3)=0.0013. 
\]

\redblock{{What do you conclude?}}

$\rightarrow$ Discussion: what  we learned from this case study. 


# What is next?

There are other results  on sampling distributions of other statistics, and  we  will cover them as needed. 
For now, look out for the use of  these theorems for:

* constructing interval estimators of unknown population  parameters, $\mu$ or $\mu_1-\mu_2$; 
* designing hypothesis tests for the same parameters; 
* a hands-on case study  and  how these are executed in R/Python.  

Note: both languages have libraries that implement these  tests and estimators for you; you don't  have to  do it from scratch. But you should know the  result that is being used, because computers don't check if theorems are applicable- they apply them when you instruct them to do so!  



#  License  

This document  is created for ITMD/ITMS/STAT 514, Spring 2021, at Illinois Tech. 

While the course materials are generally not to be distributed outside the course without permission of the instructor, all materials posted on this page are licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).
