---
title: "Topic 1 Interlude: Writing functions in R" 
author: 
  - Sonja Petrovic^[Sonja Petrović,  Associate Professor of Applied  Mathematics, College of Computing, Illinios Tech.  [Homepage](https://www.SonjaPetrovicStats.com), [Email](mailto:sonja.petrovic@iit.edu).] 
date: "Week X, ITMD/ITMS/STAT 514, Sp21"
output: rmdformats::readthedown
urlcolor: darkblue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.width = 3, fig.height = 3
  )
```


--- 

# Background 

> What is the purpose of these notes? 

1. Provide a few small examples of functions in R; 
2. Provide html/Markdown with several lines of `R` code you can use to practice writing functions. 




# Context 

How are we using R?

* Code is modular, nothing is "from scratch"
* **Functions** are the basic building block of what we do. 
 


#  Functions
- We have used a lot of built-in functions: `mean()`, `subset()`, `plot()`, `read.table()`...

- An important part of programming and data analysis is to write custom functions

- Functions help make code **modular**

- Functions make debugging easier

- Remember: this entire class is about applying *functions* to *data*

# What is a function?

> A \attn{function} is a machine that turns **input objects** (arguments) into an **output object** (return value) according to a definite rule.

- Let's look at a really simple function

```{r}
addOne <- function(x) {
  x + 1
}
```

- `x` is the **argument** or **input**

- The function **output** is the input `x` incremented by 1

```{r}
addOne(12)
```

# More interesting example

- Here's a function that returns a % given a numerator, denominator, and desired number of decimal values

```{r}
calculatePercentage <- function(x, y, d) {
  decimal <- x / y# Calculate decimal value
  round(100 * decimal, d) # Convert to % and round to d digits
}

calculatePercentage(27, 80, 1)
```

- If you're calculating several %'s for your report, you should use this kind of function instead of repeatedly copying and pasting code

# Function returning a list
- Here's a function that takes a person's full name (FirstName LastName), weight in lb and height in inches and converts it into a list with the person's first name, person's last name, weight in kg, height in m, and BMI.
\small
```{r}
createPatientRecord <- function(full.name, weight, height) {
  name.list <- strsplit(full.name, split=" ")[[1]]
  first.name <- name.list[1]
  last.name <- name.list[2]
  weight.in.kg <- weight / 2.2
  height.in.m <- height * 0.0254
  bmi <- weight.in.kg / (height.in.m ^ 2)
  list(first.name=first.name, last.name=last.name, 
       weight=weight.in.kg, height=height.in.m,
       bmi=bmi)
}
```
 \normalsize 
 
# Trying out the function
```{r}
createPatientRecord("Michael Smith", 185, 12 * 6 + 1)
```

# Another example: 3 number summary
- Calculate mean, median and standard deviation
```{r}
threeNumberSummary <- function(x) {
  c(mean=mean(x), median=median(x), sd=sd(x))
}
x <- rnorm(100, mean=5, sd=2) # Vector of 100 normals 
                              # with mean 5 and sd 2
threeNumberSummary(x)
```


# If-else statements

- Oftentimes we want our code to have different effects depending on the features of the input

- Example: Calculating a student's letter grade
  - If grade >= 90, assign A
  - Otherwise, if grade >= 80, assign B
  - Otherwise, if grade >= 70, assign C
  - In all other cases, assign F

- To code this up, we use if-else statements

# If-else Example: Letter grades
\small 
```{r}
calculateLetterGrade <- function(x) {
  if(x >= 90) {
    grade <- "A"
  } else if(x >= 80) {
    grade <- "B"
  } else if(x >= 70) {
    grade <- "C"
  } else {
    grade <- "F"
  }
  grade
}

course.grades <- c(92, 78, 87, 91, 62)
sapply(course.grades, FUN=calculateLetterGrade)
```

# `return()`

- In the previous examples we specified the output simply by writing the output variable as the last line of the function

- More explicitly, we can use the `return()` function

```{r}
addOne <- function(x) {
  return(x + 1)
}

addOne(12)
```

- We will generally avoid the `return()` function, but you can use it if necessary or if it makes writing a particular function easier.
- Google's style guide suggests explicit returns.  Most do not. 

# License

This document  is created for Math 563, Spring 2021, at Illinois Tech. 
While the course materials are generally not  to be distributed outside the course without permission of the instructor,  this particular set of notes is licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).


[![Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License][image]][hyperlink]

  [hyperlink]: https://creativecommons.org/licenses/by-nc-sa/4.0/
  [image]: https://i.creativecommons.org/l/by-nc-sa/4.0/88x31.png
    
This set of slides is taken from Prof. Alexandra Chouldechova at [CMU](https://www.andrew.cmu.edu/user/achoulde/94842/), under the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International [License](https://creativecommons.org/licenses/by-nc-sa/4.0/).
