---
title: "Topic 1 Interlude: R - Basic cleaning, loops, and alternatives" 
author: 
  - Sonja Petrovic^[Sonja Petrović,  Associate Professor of Applied  Mathematics, College of Computing, Illinios Tech.  [Homepage](https://www.SonjaPetrovicStats.com), [Email](mailto:sonja.petrovic@iit.edu).] 
date: "Week 5, ITMD/ITMS/STAT 514, Sp21"
output: rmdformats::readthedown
urlcolor: darkblue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,comment = '', fig.align = 'center' , out.width="50%"
)
```



--- 

# Context 

> What is the purpose of these notes? 

1. Provide a few small examples of functions in R; 
2. Provide html/Markdown with several lines of `R` code you can use to practice writing functions. 

## Agenda

- A common data cleaning task
- For/while loops to iterate over data
- Helpful variants of `map`, `mutate` and `summarize`

## Package loading
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
Cars93 <- MASS::Cars93  # For Cars93 data again
```

# A common problem: messy data

- One of the most common problems you'll encounter when importing manually-entered data is inconsistent data types within columns

- For a simple example, let's look at `TVhours` column in a messy version of the survey data from the following link  <http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data2019_messy.csv>: 

 
```{r, size="tiny"}
survey.messy <- read.csv("http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data2019_messy.csv", 
                         header=TRUE)
survey.messy$TVhours
```

## What's happening?
 

```{r}
str(survey.messy)
```

- Several of the entries have non-numeric values in them (they contain strings)

- As a result, `TVhours` is being imported as `factor`

## A look at the TVhours column

```{r}
survey.messy$TVhours
```

## Attempt at a fix
- What if we just try to cast it back to numeric?

```{r}
tv.hours.messy <- survey.messy$TVhours
tv.hours.messy
as.numeric(tv.hours.messy)
```

## That didn't work...
```{r, eval = FALSE}
tv.hours.messy
as.numeric(tv.hours.messy)
```
```{r, echo = FALSE}
# Show this output, but not this code
head(tv.hours.messy, 40)
head(as.numeric(tv.hours.messy), 40)
```



- This just converted all the values into the integer-coded levels of the factor

- Not what we wanted!

## Something that does work

- Consider the following simple example

```{r}
num.vec <- c(3.1, 2.5)
as.factor(num.vec)
as.numeric(as.factor(num.vec))
as.numeric(as.character(as.factor(num.vec)))
```

> If we take a number that's being coded as a factor and *first* turn it into a `character` string, *then* converting the string to a numeric gets back the number

## Back to the corrupted TVhours column

```{r}
as.character(tv.hours.messy)
as.numeric(as.character(tv.hours.messy))
typeof(as.numeric(as.character(tv.hours.messy)))  # Success!! (Almost...)
```

## A small improvement

- All the corrupted cells now appear as `NA`, which is R's missing indicator

- We can do a little better by cleaning up the vector once we get it to `character` form

```{r}
tv.hours.strings <- as.character(tv.hours.messy)
tv.hours.strings
```

## Deleting non-numeric (or .) characters 
```{r}
tv.hours.strings
# Use gsub() to replace everything except digits and '.' with a blank ""
gsub("[^0-9.]", "", tv.hours.strings) 
```

## The final product
```{r}
tv.hours.messy[1:30]
tv.hours.clean <- as.numeric(gsub("[^0-9.]", "", tv.hours.strings))
tv.hours.clean
```
- As a last step, we should go through and figure out if any of the `NA` values should really be `0`.  
  - This step is not shown here.

## Rebuilding our data
```{r}
survey <- mutate(survey.messy, TVhours = tv.hours.clean)
str(survey)
```

- **Success!**

## A different approach

- We can also handle this problem by setting `stringsAsFactors = FALSE` when importing our data.

```{r}
survey.meayssy <- read.csv("http://www.andrew.cmu.edu/user/achoulde/94842/data/survey_data2019_messy.csv", 
                         header=TRUE, stringsAsFactors=FALSE)
str(survey.messy)

```

- Now everything is a `character` instead of a `factor`

## One-line cleanup 

- Let's clean up the `TVhours` column and cast it to numeric all in one command

```{r}
survey <- mutate(survey.messy, 
        TVhours = as.numeric(gsub("[^0-9.]", "", TVhours)))
str(survey)
```

## What about all those other `character` variables?

```{r}
table(survey[["Program"]])
table(as.factor(survey[["Program"]]))
```

- Having factors coded as characters may be OK for many parts of our analysis

## If we wanted to, here's how we could fix things

`mutate_if(.tbl, .predicate, .funs)` applies a functions `.funs` to the elements of `.tbl` for which the predicate (condition) `.predicate` holds.

Here is how we can use `mutate_if` to convert all character columns to factors.
```{r}
survey <- survey %>% mutate_if(is.character, as.factor)
# Outcome:
str(survey)
```

- **Success!**

# Another common problem

- In various homework assignments, you'll learn how to wrangle with another common problem

- When data is entered manually, misspellings and case changes are very common

- E.g., a column showing treatment information may look like,

```{r}
treatment <- as.factor(c("dialysis", "Ventilation", "Dialysis", "dialysis", "none", "None", "nnone", "dyalysis", "dialysis", "ventilation", "none"))
summary(treatment)
```
# 
```{r}
summary(treatment)
```

- This factor has 8 levels even though it should have 3 (dialysis, ventilation, none)

- We can fix many of the typos by running spellcheck in Excel before importing data, or by changing the values on a case-by-case basis later

- There's a faster way to fix just the capitalization issue (this will be an exercise in one of the HW sets)


# What are all these `map<*>` functions?

- These are all efficient ways of applying a function to margins of an array or elements of a list

- Before we talk about them in detail, we need to understand their more cumbersome but more general alternative: loops

- **loops** are ways of iterating over data

- The `map<*>` functions and their `<*>apply` base-R ancestors can be thought of as good *alternatives* to loops 

# `for` loops

## For loops: a pair of examples
```{r}
for(i in 1:4) {
  print(i)
}

phrase <- "Good Night,"
for(word in c("and", "Good", "Luck")) {
  phrase <- paste(phrase, word)
  print(phrase)
}
```

## For loops: syntax

> A **for loop** executes a chunk of code for every value of an **index variable** in an **index set**

- The basic syntax takes the form

```{r, eval=FALSE}
for(index.variable in index.set) {
  code to be repeated at every value of index.variable
}
```

- The index set is often a vector of integers, but can be more general

## Example
```{r}
index.set <- list(name="Michael", weight=185, is.male=TRUE) # a list
for(i in index.set) {
  print(c(i, typeof(i)))
}
```

## Example: Calculate sum of each column
```{r}
fake.data <- matrix(rnorm(500), ncol=5) # create fake 100 x 5 data set
head(fake.data,2) # print first two rows

col.sums <- numeric(ncol(fake.data)) # variable to store running column sums
for(i in 1:nrow(fake.data)) {
  col.sums <- col.sums + fake.data[i,] # add ith observation to the sum
}
col.sums

colSums(fake.data) # A better approach (see also colMeans())
```

# `while` loops

- **while loops** repeat a chunk of code while the specified condition remains true

```{r, eval=FALSE}
day <- 1
num.days <- 365
while(day <= num.days) {
  day <- day + 1
}
```

- We won't really be using while loops in this class

- Just be aware that they exist, and that they may become useful to you at some point in your analytics career

# Loop alternatives

 

Command | Description
--------|------------
`apply(X, MARGIN, FUN)` | Obtain a vector/array/list by applying `FUN` along the specified `MARGIN` of an array or matrix `X`
`map(.x, .f, ...)` | Obtain a *list* by applying `.f` to every element of a list or atomic vector `.x`
`map_<type>(.x, .f, ...)` | For `<type>` given by `lgl` (logical), `int` (integer), `dbl` (double) or `chr` (character), return a *vector* of this type obtained by applying `.f` to each element of `.x` 
`map_at(.x, .at, .f)` | Obtain a *list* by applying  `.f` to the elements of `.x` specified by name or index given in `.at`
`map_if(.x, .p, .f)` | Obtain a *list* `.f` to the elements of `.x` specified by `.p` (a predicate function, or a logical vector)
`mutate_all/_at/_if` | Mutate all variables, specified (at) variables, or those selected by a predicate (if)
`summarize_all/_at/_if` | Summarize all variables, specified variables, or those selected by a predicate (if)

- These take practice to get used to, but make analysis easier to debug and less prone to error when used effectively

- The best way to learn them is by looking at a bunch of examples.  The end of each help file contains some examples.  

## Example: apply()
```{r}
colMeans(fake.data)
apply(fake.data, MARGIN=2, FUN=mean) 
# MARGIN = 1 for rows, 2 for columns
# Function that calculates proportion
# of vector indices that are > 0
propPositive <- function(x) mean(x > 0)
apply(fake.data, MARGIN=2, FUN=propPositive) 
```

## Example: map, map_<type>()
```{r}
map(survey, is.factor) # Returns a list
map_lgl(survey, is.factor) # Returns a logical vector with named elements
```

## Example: apply(), map(), map_<type>()
```{r}
apply(cars, 2, FUN=mean) # Data frames are arrays
map(cars, mean) # Data frames are also lists
map_dbl(cars, mean) # map output as a double vector
```

## Example: mutate_if

Let's convert all factor variables in Cars93 to lowercase


```{r}
head(Cars93$Type)
Cars93.lower <- mutate_if(Cars93, is.factor, tolower)
head(Cars93.lower$Type)
```
Note: this operation is performed "in place", replacing prior versions of the variables

## Example: mutate_if, adding instead of replacing columns

If you pass the functions in as a list with named elements, those names get appended to create modified versions of variables instead of replacing existing variables

```{r}
Cars93.lower <- mutate_if(Cars93, is.factor, 
                          list(lower = tolower))
head(Cars93.lower$Type)
head(Cars93.lower$Type_lower)
```

## Example: mutate_at

Let's convert from MPG to KPML but this time using `mutate_at`
```{r}
Cars93.metric <- Cars93 %>% 
  mutate_at(c("MPG.city", "MPG.highway"), 
            list(KMPL = ~ 0.425 * .x))
tail(colnames(Cars93.metric))
```
Here, `~ 0.425 * .x` is an example of specifying a "lambda" (anonymous) function.  It is permitted short-hand for 

```{r, eval = FALSE}
function(.x){0.425 * .x}
```


## Example: summarize_if

Let's get the mean of every numeric column in Cars93
\tiny
 
```{r}
Cars93 %>% summarize_if(is.numeric, mean)

Cars93 %>% summarize_if(is.numeric, list(mean = mean), na.rm=TRUE)
```

## Example: summarize_at

Let's get the average fuel economy of all vehicles, grouped by their Type

We'll use the helper function `contains()` to indicate that we want all variables that contain MPG.

```{r}
Cars93 %>%
  group_by(Type) %>%
  summarize_at(c("MPG.city", "MPG.highway"), mean)
```

# Another approach

We might  learn about a bunch of helper functions like `contains()` and `starts_with()`.  Here's one way of performing the previous operation with the help of these functions, and appending `_mean` to the resulting output. 

```{r}
Cars93 %>%
  group_by(Type) %>%
  summarize_at(vars(contains("MPG")), list(mean = mean))
```

# More than one grouping variable

```{r}
Cars93 %>%
  group_by(Origin, AirBags) %>%
  summarize_at(vars(contains("MPG")), list(mean = mean))
```


# License

This document  is created for Math 514, Spring 2021, at Illinois Tech. 
While the course materials are generally not  to be distributed outside the course without permission of the instructor,  this particular set of notes is licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).


[![Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License][image]][hyperlink]

  [hyperlink]: https://creativecommons.org/licenses/by-nc-sa/4.0/
  [image]: https://i.creativecommons.org/l/by-nc-sa/4.0/88x31.png
    
This set of notes is adapted from Prof. Alexandra Chouldechova at 
CMU, under the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International [License](https://creativecommons.org/licenses/by-nc-sa/4.0/).

